\Huge Conferences in an Era of Expensive Carbon

### 2019-06-10 mike.tex
Comments on Mike proposal, and some polishing on it

A broad scientific consensus warns us that human
emissions of carbon dioxide and other greenhouse gases are warming the earth.
% (Figure \ref{fig:warming}).
This is not a problem to be left to future
generations: we are up against it now. The UN's Intergovernmental
Panel on Climate Change (IPCC) says a 40\% decrease in
emissions by 2030 is needed to avoid irreversible damage~\cite{IPCC-report}.
% (Figure~\ref{fig:IPCC-cover}).
Achieving reductions on this scale will
require urgent and sustained commitment at all levels of society, including
large organizations like universities, companies, and scientific societies
like ACM.

Indeed, scientific societies arguably have an especially important role to
play, since one of their main activities is organizing international
conferences, which produce significant greenhouse gas emissions,
particularly from air travel. A single round-trip flight from
Philadelphia to Paris emits the equivalent of about 1.7 tons of carbon
dioxide (\COtwoE) per passenger~\cite{carboncalculator}\iflater\bcp{Trim
  citation?}\fi---a substantial fraction of the {\em total} yearly emissions
for an average resident of the US (16.5 tons) or Europe (7
tons)~\cite{emissions}.  \iflater\bcp{For comparison, ...  (put in some
  figures about the footprints of a few other things that people might think
  of reducing---driving, eating meat, upgrading their fridge...)}\fi{}
Moreover, these emissions have no near-term 
technological fix, since jet fuel is difficult to replace with
renewable energy sources~\cite{elec-air}. 
% \iflater\bcp{Trimmed this sentence: Reducing the \COtwoE{} footprint
% of conferences will require a reduction in the footprint of the attendees'
% air travel.}\fi

We formed the ACM SIGPLAN Climate Committee with this situation in
mind. We made presentations at SIGPLAN events, had discussions with
ACM and SIG leaders, and ultimately put together a report of ideas for
reducing the \COtwoE{} costs of conference
activities~\cite{ClimateCommitteReport}. The report considers easy 
but incremental changes like making program committee meetings virtual
instead of physical and enabling virtual participation via live-streaming,
all the way to complete re-imagining of conferences and their role in
science (holding conferences completely in virtual spaces, replacing
conferences with journals, etc.), with some intriguing hybrid forms in the
middle (regional conferences, mega-conferences, conferences split into
two locations on different continents, etc.).

@@

Those we spoke to were almost universally supportive of reducing
\COtwoE{} emissions due to conferences, but very few were eager
to change the way we organize them. This
hesitation makes sense. Science
is a fundamentally social process, and the conference system
accelerates research through high-bandwidth interaction, direct
dissemination of results, network building, and serendipitous
cross-fertilization. How to weigh the benefit to the environment of a
particular \COtwoE-reducing change, such as on-line PC meetings,
against the harm the change could cause to the scientific process,
such as the lost opportunity to build networks and germinate
synergistic ideas?

Scientific value and environmental harm are apples and oranges, so
they are hard to compare. They are also very hard to measure. While we
can reasonably estimate the \COtwoE{} saved by a particular change, it
is harder to guess the impact of the example it sets. This impact is
important. Emissions from air travel to conferences are the lion's
share of ACM's \COtwoE{} footprint, but air travel only contributes 
2\% of current global emissions~\cite{emissions}. This small fraction is not an excuse
for inaction. Just because any individual's or
organization's share of global emissions is miniscule does not mean it
doesn't matter. We must all contribute to the global 40\% reduction
needed by 2030 in a way that makes the most difference, and we must set an example for
others to do the same. It is especially important for scientists---who
as a community have been raising the alarm about climate change---to
not appear to be declaring themselves exempt from cutting back. But
they must do so in a smart way, or else risk disproportionately
harming the progress that science can bring.

@@

Fortunately, there is a well-established mechanism for comparing apples and
oranges---indeed, for comparing almost anything!  Economists call it ``the
magic of markets.''  By mapping choices and preferences in disparate domains
onto a common scale (money), economic markets can be viewed as massively
distributed computations that optimize costs and benefits across vastly
different activities.\bcp{Give a simple, concrete example.}  For this
reason, many policy experts advocate addressing climate change through some
form of carbon tax, which places a monetary cost on greenhouse gas
emissions, sending a clear market signal that changes aggregate
behavior.\bcp{See
  https://kleinmanenergy.upenn.edu/policy-digests/why-carbon-pricing-falls-short
  for lots of good phrasing and citations.}  Exactly how the price of carbon
should be set, and what should be done with the raised funds, is a
challenging policy question, but it has a clear goal: It 
should be chosen so that the world arrives at zero emissions within the
timeframe that climate scientists tell us is safe.

While a global \COtwoE{} market is the ultimate goal, we do not need
to get there in one fell swoop.
Some countries\bcp{e.g.?}, US states (e.g., California), universities
(e.g., UMD and UCLA), and companies (\mwh{Jane Street?}) have
begun to adopt internal carbon pricing policies and markets, to push
their own behavior closer to a sustainable equilibrium. We feel that ACM
should do the same thing. Here's how it might start.

@@

The main idea is to for ACM to enforce a carbon tax on conference
budgets. At present, ACM charges an ``allocation'' percentage of
12\% on total revenue. It can also impose a tax on
the \COtwoE{} emissions induced by the conference. This tax has two
benefits. First, it creates an incentive for SIGs and conference organizers to
lower total costs by reducing emissions, so they can pay less
tax. They could do this in many ways, such as the ones outlined in our 
report. Second, ACM can use the extra revenue for ``good works.'' Such
good works could directly offset the carbon that paid for them,
or they could provide some other benefit, such as defraying open access
publishing costs.

@@

For the first, we can start by estimating \COtwoE{} emissions due to
conference travel by analyzing the conference's registration data. We
worked the ACM SGB to develop a \COtwoE{} calculator that takes such
registration data and estimates the \COtwoE{} cost borne by each
individual's travel to the conference. It could become part of
standard practice to run this calculator when closing out the
conference, to assess the tax. \mwh{other \COtwoE{} costs?}
%
Moreover, intrepid SIG leaders and conference organizers can analyze
the gathered data to look for opportunities for reduction. For
example, we used the calculator to compute the per-capita and total
\COtwoE{} costs of several SIGPLAN conferences over the last few years.
\mwh{Discuss.}
\bcp{Put the discussion of Figure 1 (and an accompanying figure
showing distance traveled to several instances of the same conference) here;
this illustrates the difficulty and the need for careful data gathering and
analysis.}

@@

How should we set the price of a conference's \COtwoE? There are several
possibilities. One idea is simply to adopt the price from an existing
market, such as California's; between 2012 and 2018, the price of a
ton of \COtwoE{} fluctuated from about \$12 to \$23~\cite{CACO2}. Another idea is to
match the price to the equivalent in \emph{carbon
  offsets}~\cite{CarbonOFfsetReport}. Offset providers use raised funds
to support an activity (e.g., installing methane capture devices on
landfill sites or buying fuel-efficient stoves to replace open-fire
cooking in poor communities) that is {\em certified} (by yet another
organization) to permanently avoid emitting an amount of greenhouse
gases specified. Current prices vary widely around an average of around
\$15 per ton, depending on the vendor, the nature of the offset
project, the region of the world where it will be implemented, etc.

However we estimate the price, we could imagine setting it a little low at the
outset, perhaps based on analysis of existing conference \COtwoE{} usage, and then raising it
over time as organizers start to adapt to it. On the other hand, the true
``social cost'' of a ton of carbon is variously estimated to be much
higher than the above---anywhere from \$40 to \$400 per
ton~\bcp{citation}---so current pricing may already be on the low
side.

@@

The climate crisis is too large and too urgent to leave to world leaders to
address at their own pace: organizations at every scale should confront
their own contributions, raise awareness among their members, and begin
establishing new ways of doing business in the lower-carbon future that is
coming.
% For ACM, this especially means confronting the carbon-intensive
% nature of our conferences.
% %
% You can help accelerate this process in a number of ways: 
% \begin{itemize}
% \item {\bf Get involved} in raising awareness about this issue in the
% organizations you are a part of.  \bcp{Does the rest of this smack of
%   incrementalism?  Maybe delete it.  ``For example, ask the organizers of
%   the conferences you attend what they are doing about reducing or
%   offsetting emissions from participants' travel.''}
% \item {\bf Connect} with others that are engaging with these issues---in
% your workplace, in your other communities, and within ACM (e.g., on the
% {\tt acm-climate} mailing list~\cite{acm-climate}).
% \item {\bf Analyze} your organization's carbon-emitting activities to
% understand how to reduce its footprint now and prepare for the coming era of
% expensive carbon.
% \end{itemize}

>>>>

A broad scientific consensus warns us that human
emissions of carbon dioxide and other greenhouse gases are warming the earth.
% (Figure \ref{fig:warming}).
This is not a problem we can leave to future
generations: The UN's Intergovernmental
Panel on Climate Change (IPCC) says a 40\% decrease in
emissions by 2030 is needed to avoid irreversible damage~\cite{IPCC-report}.
% (Figure~\ref{fig:IPCC-cover}).
Achieving reductions on this scale will
require urgent and sustained commitment at all levels of society, including
not only city, state, and national governments, but also 
universities, companies, and scientific societies
like ACM.

Indeed, scientific societies have an especially important role to
play, since one of their main activities is organizing international
conferences, which produce significant greenhouse gas emissions,
particularly from air travel. A single round-trip flight from
Philadelphia to Paris emits the equivalent of about 1.7 tons of carbon
dioxide (\COtwoE) per passenger~\cite{carboncalculator}\iflater\bcp{Trim
  citation?}\fi---a substantial fraction of the {\em total} yearly emissions
for an average resident of the US (16.5 tons) or Europe (7
tons)~\cite{emissions}.  \iflater\bcp{For comparison, ...  (put in some
  figures about the footprints of a few other things that people might think
  of reducing---driving, eating meat, upgrading their fridge...)}\fi{}
Moreover, these emissions have no near-term 
technological fix, jet fuel being difficult to replace with
renewable energy sources~\cite{elec-air}. 
% \iflater\bcp{Trimmed this sentence: Reducing the \COtwoE{} footprint
% of conferences will require a reduction in the footprint of the attendees'
% air travel.}\fi

Motivated by this situation, in 2016 we formed the ACM SIGPLAN Climate
Committee. Since then, we made presentations at SIGPLAN events, spoken with
ACM and SIG leaders, and assembled a report assessing a number of ideas for
reducing the \COtwoE{} costs of conference
activities~\cite{ClimateCommitteReport}.  Our report considers a range of
options, fromn easy-but-incremental changes like making program committee
meetings virtual instead of physical and enabling virtual participation in
conferences via live-streaming, all the way to complete
re-conceptualizations of conferences and their role in science (e.g.,
replacing conferences with journals, holding conferences completely in
virtual spaces, etc.), as well as some intriguing points in the middle of
the spectrum (regional conferences, mega-conferences, conferences held
simultaneously at two locations on different continents, etc.).

@@

Those we spoke to have been almost universally supportive of reducing
\COtwoE{} emissions due to conferences, but very few are eager to
fundamentally change the way conferences are organized. This hesitation
makes sense. Science is a fundamentally social process, and the conference
system accelerates research through high-bandwidth interaction, direct
dissemination of results, network building, and serendipitous
cross-fertilization.  So how can we weight the ennvironmental benefit of a
particular \COtwoE-reducing change, such as on-line PC meetings, against the
harm the change could cause to the scientific process, such as the lost
opportunity to build networks and germinate synergistic ideas?

Scientific value and environmental harm are apples and oranges---hard to
compare.  Indeed, they are even hard to measure. In particular, while we can
reasonably estimate the \COtwoE{} directly saved by a particular change, it
is much harder to assess the impact of the example it sets.

This impact is important---perhaps more important than \COtwoE{} reductions
themselves. Emissions from air travel to conferences are the lion's share of
ACM's \COtwoE{} footprint, but air travel only contributes 2\% of current
global emissions~\cite{emissions}. But this small fraction should not be
taken as an excuse for inaction: the fact that any individual's or
organization's share of global emissions is miniscule does not mean it
doesn't matter. {\em No} individual's (or organization's) contributions
amount to a nontrivial percentage of global emissions; {\em everyone} will
need to contribute, to achieve the global 40\% reduction needed by 2030. To
get others on board, we must set an example. Indeed, it is especially
important for scientists---who, as a community, have been instrumental in
raising the alarm about climate change---to not appear to be exempting
themselves from cutting back.  But they must do so in a smart way, or else
risk disproportionately harming the progress that science can bring.

@@

Fortunately, there is a well-established mechanism for comparing apples and
oranges---indeed, for comparing almost anything!  Economists call it ``the
magic of markets.''  By mapping choices and preferences in disparate domains
onto a common scale (i.e., money), economic markets function as
massively distributed computations that optimize costs and benefits across
vastly different activities.\bcp{Give a simple, concrete example.}  For this
reason, many policy experts advocate addressing climate change through some
form of carbon pricing, which places a monetary cost on greenhouse gas
emissions, sending a clear market signal that changes aggregate
behavior.\bcp{See
  https://kleinmanenergy.upenn.edu/policy-digests/why-carbon-pricing-falls-short
  for lots of good phrasing and citations.}  Exactly how the price of carbon
should be set, and what should be done with the raised funds, is a
challenging policy question, but it has a clear goal: It should be chosen so
that the world arrives at zero emissions within the timeframe that climate
scientists tell us is safe.

While a global \COtwoE{} market is the ultimate goal, we do not need
to get there in one fell swoop.
Some countries\bcp{e.g.?}, US states (e.g., California), universities
(e.g., UMD and UCLA), and companies (\mwh{Jane Street?}) have already
begun to adopt internal carbon pricing policies and markets, to push
their own behavior closer to a sustainable equilibrium.  We argue that ACM
should do the same thing. Here's how it might start.

@@

The main idea is to for ACM to enforce a carbon tax on conference
budgets. At present, ACM charges an ``allocation'' percentage of
12\% on total revenue. It can also impose a tax on
the \COtwoE{} emissions induced by the conference. This tax has two
benefits. First, it creates an incentive for SIGs and conference organizers to
lower total costs by reducing emissions, so they can pay less
tax. They could do this in many ways, such as the ones outlined in our 
report. Second, ACM can use the extra revenue for ``good works.'' Such
good works could directly offset the carbon that paid for them,
or they could provide some other benefit, such as defraying open access
publishing costs.  \bcp{I foresee a tricky legal question of whether it's OK for
  ACM to charge such a tax simply for ``good works.''}

@@

For the first, we can start by estimating \COtwoE{} emissions due to
conference travel by analyzing the conference's registration data. We
worked with the ACM SGB to develop a \COtwoE{} calculator that takes such
registration data and estimates the \COtwoE{} cost borne by each
individual's travel to the conference. It could become part of
standard practice to run this calculator when closing out the
conference, to assess the tax. \mwh{other \COtwoE{} costs?}  \bcp{But the
  conference planners would need at least an estimate of what the tax was
  going to amount to, to avoid a budget shortfall.}
%
Moreover, intrepid SIG leaders and conference organizers can analyze
the gathered data to look for opportunities for reduction. For
example, we used the calculator to compute the per-capita and total
\COtwoE{} costs of several SIGPLAN conferences over the last few years.
\mwh{Discuss.}
\bcp{Put the discussion of Figure 1 (and an accompanying figure
showing distance traveled to several instances of the same conference) here;
this illustrates the difficulty and the need for careful data gathering and
analysis.}

@@

How should we set the price of a conference's \COtwoE? There are several
possibilities. One idea is simply to adopt the price from an existing
market, such as California's; between 2012 and 2018, the price of a
ton of \COtwoE{} fluctuated from about \$12 to \$23~\cite{CACO2}. Another idea is to
match the price to the equivalent in \emph{carbon
  offsets}~\cite{CarbonOFfsetReport}. Offset providers use raised funds
to support an activity (e.g., installing methane capture devices on
landfill sites or buying fuel-efficient stoves to replace open-fire
cooking in poor communities) that is {\em certified} (by yet another
organization) to permanently avoid emitting an amount of greenhouse
gases specified. Current prices vary widely around an average of around
\$15 per ton, depending on the vendor, the nature of the offset
project, the region of the world where it will be implemented, etc.
On the other hand, the true
``social cost'' of a ton of carbon is variously estimated to be much
higher than the above---anywhere from \$40 to \$400 per
ton~\bcp{citation}---so current pricing may already be on the low
side.

However we estimate the price, we could imagine setting it a little low at the
outset, perhaps based on analysis of existing conference \COtwoE{} usage, and then raising it
over time to give organizers time to adapt.

@@

The climate crisis is too large and too urgent to leave to world leaders to
address at their own pace: organizations at every scale must confront
their own contributions, raise awareness among their members, and begin
establishing new ways of doing business in the lower-carbon future that is
coming.
\bcp{A sentence restating our proposal.}

% For ACM, this especially means confronting the carbon-intensive
% nature of our conferences.
% %
% You can help accelerate this process in a number of ways: 
% \begin{itemize}
% \item {\bf Get involved} in raising awareness about this issue in the
% organizations you are a part of.  \bcp{Does the rest of this smack of
%   incrementalism?  Maybe delete it.  ``For example, ask the organizers of
%   the conferences you attend what they are doing about reducing or
%   offsetting emissions from participants' travel.''}
% \item {\bf Connect} with others that are engaging with these issues---in
% your workplace, in your other communities, and within ACM (e.g., on the
% {\tt acm-climate} mailing list~\cite{acm-climate}).
% \item {\bf Analyze} your organization's carbon-emitting activities to
% understand how to reduce its footprint now and prepare for the coming era of
% expensive carbon.
% \end{itemize}
